---
title: "final_exam"
author: "Rick Shemanski"
date: "March 17, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r librarys, echo=FALSE, message=F}
library(dplyr)
library(nycflights13)
library(RSQLite)
library(ggplot2)
flights_sqlite <- tbl(nycflights13_sqlite(), "flights")
planes_sq <-tbl(nycflights13_sqlite(), "planes")
weather_sq <- tbl(nycflights13_sqlite(), "weather")

```



```{r canceled flights, echo = FALSE, message=F}
f00 <-   flights_sqlite%>%
  mutate(
    canceled = is.na(arr_time)
  ) 

tar_orig <- c("EWR", "JFK", "LGA")
f1 <- filter(f00, year == 2013, canceled ==1)
f11 <- filter(f1, origin %in% tar_orig)
f2 <- select(f11, year, month, day, hour, minute, origin, dest, distance, 
             dep_time, dep_delay, canceled, carrier, tailnum, flight)

```


```{r planes, echo= F}
p0 <- mutate(planes_sq, 
             year_made = year
             )%>%
  select(-year)
planes_df <- collect(p0)

```



```{r join_planes, echo=F }
## join in sql and then assign to flights_planes_test

## f2 is cancelled flights only- 
flights_planes_canceled <- left_join(
  f2, p0, 
  by = "tailnum") %>%
  collect() 

## create date for merged data- planes and flights

flights_planes_canceled <- flights_planes_canceled%>%
  mutate(
      date = paste(year, month, day, sep ="-"),
      date = as.Date(date, format = "%Y-%m-%d")
  )%>%
  select(-year)


```





```{r weather_preprocess, echo=F}

w1 <- filter(weather_sq, year == 2013)
weather_df <- collect(w1)

weather_df <- weather_df%>%
  mutate(
    date = paste(year, month, day, sep ="-"),
    date = as.Date(date, format = "%Y-%m-%d")
  )
#typeof(weather_test$date) --- good

## collapse to avererage daily weather- 
by_origin <- group_by(weather_df, origin,date)


weather_mean <- by_origin%>%
  summarise(
    #origin = median(origin, na.rm = T)
    temp = mean(temp, na.rm=T), 
    dewp = mean(dewp,na.rm=T),
    humid = mean(humid,na.rm=T),
    wind_dir= mean(wind_dir,na.rm=T),
    wind_speed = mean(wind_speed,na.rm=T),
    wind_gust= mean(wind_gust,na.rm=T),
    precip = mean(precip,na.rm=T),
    pressure = mean(pressure,na.rm=T),
    visib = mean(visib,na.rm=T)
  )

```



```{r final_df_canc ,echo = FALSE}

### daily  weather, plane description , canceled flights
final_df_canceled <- left_join(flights_planes_canceled, weather_mean, by = c("origin"= "origin", "date" = "date")) 

```

Cancellations vs Months  by origin

```{r plots, echo= FALSE}

p <- ggplot(final_df_canceled, aes(x = month, fill = origin))
p <- p + geom_histogram(binwidth = 0.5)
p + ggtitle("Cancelations by Month")

p3 <- ggplot(final_df_canceled, aes(factor(month), fill = factor(origin))) + geom_bar(position = "dodge")
p3 + ggtitle("Cancelations by Month by Aiport")

```


The above plots show the cancellations by month for NYC airport. This information can help travel choose a originating airport based off what time of year they are traveling.  These numbers are in total cancellations, not as a cancellation/total flight ratio. In the future a ratio will be used to ensure that origins with greater total flights are not misrepresented. 


```{r top_canceled_dest, echo=FALSE}
###originality !!!!!!
top_dest_Delays <- group_by(final_df_canceled, dest, month)%>%
  summarise(
    canceled_flights = sum(canceled)
  )

top_dest_Delays <- group_by(top_dest_Delays,month)%>%
  mutate(
    month_85 = quantile(canceled_flights,prob= c(0.85), na.rm= TRUE),
    top_85 = ifelse(canceled_flights > month_85, 1, 0)
  )

top_worst <- filter(top_dest_Delays, top_85 == 1)

top_worst1 <- filter(top_worst, month == 1)
top_worst4 <- filter(top_worst, month == 4)
top_worst8 <- filter(top_worst, month == 8)
top_worst12 <- filter(top_worst, month == 12)


```


```{r worst_dest_plots, echo=FALSE}

p_1 <- ggplot(top_worst1, aes(x= dest, y = canceled_flights, fill= dest))
p_1<- p_1 + geom_bar(stat= "identity")

p_1 <- p_1 + ggtitle("Top 20% of Canceled Destinations Month 1")


p_4 <- ggplot(top_worst4, aes(x= dest, y = canceled_flights, fill= dest))
p_4<- p_4 + geom_bar(stat= "identity")

p_4 <- p_4 + ggtitle("Top 20% of Canceled Destinations Month 4")

p_8 <- ggplot(top_worst8, aes(x= dest, y = canceled_flights, fill= dest))
p_8<- p_8 + geom_bar(stat= "identity")

p_8 <- p_8 + ggtitle("Top 20% of Canceled Destinations Month 8")


p_12 <- ggplot(top_worst12, aes(x= dest, y = canceled_flights, fill= dest))
p_12 <- p_12 + geom_bar(stat= "identity")

p_12 <- p_12 + ggtitle("Top 20% of Canceled Destinations Month 12")

```

Cancellations vs Destination

```{r worst_dest_plot_print, echo=FALSE}
p_1
p_4
p_8
p_12
```

Below we examined the cancellations for each destination for a specific month. We compared the cancellations at a specific destination to all other destinations' cancellations. We then determined which destinations were the Top 20% Most Canceled Destinations for a specific month. In the future we also need to use a cancellation ratio, so to account for larger airports (ATL, DALLAS, etc...) having a greater number of cancellations. 

This section I feel would be very different than anyone else. Code starts on RMD135 





```{r delays_all, echo=FALSE}
f00 <-   flights_sqlite%>%
  mutate(
    canceled = is.na(arr_time)
  ) 

tar_orig <- c("EWR", "JFK", "LGA")
f1 <- filter(f00, year == 2013)
f11 <- filter(f1, origin %in% tar_orig)
f2 <- select(f11, year, month, day, hour, minute, origin, dest, distance, dep_time, dep_delay, canceled, carrier, tailnum, flight)

```




```{r new_data, echo=FALSE}

## planes preprocessing

##planes -- change year to year made- production year 
p0 <- mutate(planes_sq, 
             year_made = year
)%>%
  select(-year)
planes_df <- collect(p0)


## any repeats?
length(unique(planes_df$tailnum))
#######################################################3333

## join in sql and then assign to flights_planes_test

## f2 is cancelled flights only- 
flights_planes_all <- left_join(
  f2, p0, 
  by = "tailnum") %>%
  collect() 


########################333


## create date for merged data- planes and flights

flights_planes_all <- flights_planes_all%>%
  mutate(
    date = paste(year, month, day, sep ="-"),
    date = as.Date(date, format = "%Y-%m-%d")
  )%>%
  select(-year)


######
######################################

w1 <- filter(weather_sq, year == 2013)
weather_df <- collect(w1)

weather_df <- weather_df%>%
  mutate(
    date = paste(year, month, day, sep ="-"),
    date = as.Date(date, format = "%Y-%m-%d")
  )
#typeof(weather_test$date) --- good

## collapse to avererage daily weather- 
by_origin <- group_by(weather_df, origin,date)


weather_mean <- by_origin%>%
  summarise(
    #origin = median(origin, na.rm = T)
    temp = mean(temp, na.rm=T), 
    dewp = mean(dewp,na.rm=T),
    humid = mean(humid,na.rm=T),
    wind_dir= mean(wind_dir,na.rm=T),
    wind_speed = mean(wind_speed,na.rm=T),
    wind_gust= mean(wind_gust,na.rm=T),
    precip = mean(precip,na.rm=T),
    pressure = mean(pressure,na.rm=T),
    visib = mean(visib,na.rm=T)
  )

### works --
testing <- filter(weather_df, year ==2013, month==1, day ==1)
mean_test<- mean(testing$temp, na.rm = T)
#mean_test


### join canceled planes and flights to the daily weather for each origin. 

final_all <- inner_join(flights_planes_all, weather_df, by = c("origin"= "origin", "date" = "date", "hour"="hour")) 
final_all$delayed <- ifelse(final_all$dep_delay > 10 , 1, 0)


```


```{r weather_regressions, echo=FALSE}

## weather plotting
weather_df1 <- final_all%>%
  filter(origin== "EWR")

log_weather <- glm(canceled~ hour+temp + wind_speed + wind_gust+ visib + precip + pressure+ humid, data = weather_df1, family= binomial(link="logit"))

##check which variables are important 
reg_weather <- glm(dep_delay~ hour+ temp + wind_speed + wind_gust+ visib + precip + pressure+ humid, data = weather_df1, family= gaussian)




```

```{r time_delay_plots, echo=FALSE}

#### time delays

## by day

we1 <- ggplot(final_all, aes(x= day.x, y= delayed))
we2 <- we1 + geom_bar(stat = "identity", aes(fill= day.x))
we2 <- we2 + ggtitle("Departure Delays by day")

## by hour
hour1 <- ggplot(final_all, aes(x= hour, y= delayed))
hour1  <- hour1  + geom_bar(stat = "identity", aes(fill= hour))
hour1 <- hour1  + ggtitle("Departure Delays by day")

```

Delays vs Time


```{r time_delay_plots_print, echo=FALSE}
we2
hour1
```

In the above plots we can see the number of delayed flights as function of time. There seems to be no strong correlation between if a flight will be delayed based upon how late the departure is into the month.

However, there seems to be a strong correlation between departure delays and the hour in which the planes depart. The later a plane departs, the more likely it will be delayed. 


```{r plane_charterisrics, echo=FALSE}

plane_char <- select(final_all, year_made, dep_delay, type)
plane_char <- na.omit(plane_char)
pl <- ggplot(plane_char, aes(x = year_made, y= dep_delay, fill = type))
pl <- pl+ geom_point(aes(color= type))
pl<- pl + facet_grid(type~.) + ggtitle("Departure Delays by Age of Plane and Plane Type")

```

Delays vs Plane Characteristics

```{r plane_char_plots, echo=FALSE}
pl
```

The above plot shows how the age of the plane, and type of plane contribute to departure delays. Most planes are fixed wing multi engine, these planes are happen to be newer than the other types. However, these newer planes still are equally delayed than order planes.

```{r weather_plots, echo=FALSE}

wind_df <- filter(final_all, wind_speed < 70)
wind_lm <- lm(dep_delay~wind_speed, data = wind_df)



wind <- ggplot(wind_df, aes(wind_speed, dep_delay))
wind_p <- wind + geom_point(aes(size= type, color = type))
wind_p <- wind_p + stat_smooth(method= lm, aes(color= type))
wind_p <- wind_p + ggtitle("Deperature Delay Vs Wind Speed by plane type")


wind_b <- ggplot(wind_df, aes(visib, canceled, fill = visib))
wind_b <- wind_b + geom_bar(stat = "identity",aes(fill= factor(visib)))
wind_b <- wind_b + ggtitle("Flights Canceled by Visibility")

```


Delays vs Weather

Wind vs Delays

```{r weather_delay_plots, echo=FALSE}
wind_p

```


In the above plot we can see how wind-speed effects departure delays by plane type. We might think that newer planes would be able to operate in higher wind-speeds, but there does not seem to be a strong indication. A linear regression line for each plane type shows how wind-speed affects departure delays. 

Visibility vs Cancelled Flights

```{r weather_delay_visibility, echo=FALSE}

wind_b
```

In the above chart we can see the number cancellations for each factor of visibility. Clearly a visibility score of 10 has the greatest number of departures. 

